---
title: "Data Education Workshop 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Introduction
The Center for Systems Science and Engineering at Johns Hopkins was among the earliest to collect and publish global data related to the COVID-19 pandemic. The data used for their [reporting dashboard] (https://coronavirus.jhu.edu/map.html), aggregated from multiple sources and updated daily, is freely available on [Github] (https://github.com/CSSEGISandData/COVID-19). 

We will review a basic overview of the following steps in data analysis: 
1. Data Import and Cleaning
2. Exploratory Data Analysis 
3. Visualization 
4. Regression and Classification

####Setting Up R: 
We will go over this in more detail in our next workshop which will help you get oriented and set up! 
1. Create New Project (Via Github Repository) & Establish Working Directory 
2. Create a File (RMD)
3. Load Packages Chunk

####Load Packages:
```{r}
library(tidyverse)
library(lubridate)
library(zoo)
library(maps)
library(tidyr)
library(ggplot2)
library(ggthemes)
```

####Data Import and Cleaning
Step 0: Finding Data (Look for CSV Files!) 
Step 1: Read in data for  (Make sure you either use raw data link if available or import your CSV file into the same working directory as your Rmd file.) 
Step 2: View Data & Decide What Needs to be "Cleaned"  
Step 3: Data Cleaning! 

```{r}
cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
``` 

Reshape the data to long format, so that instead of having many columns of cases corresponding to a different date, you have one column that stores the cases and one column that stores the dates. 

```{r}
cases_long <- cases %>% gather(date, cases, "1/22/20":"10/31/20") 
```

Convert the new date column to be a date type instead of a character type. 

```{r}
#Check Class
class(cases_long$date)

cases_long$date <-as.Date(cases_long$date, "%m/%d/%y")

#Re-Check Class
class(cases_long$date)
```

This dataset keeps track of the number of confirmed cases reported for states, provinces, and other sub-regions within a given country but we just want total country data. This data frame also still includes a bunch of other information we're not interested in (Latitude Longitude Etc...) 

We can create a new data frame with only what we're interested in: country name, date, and the total number of cases in the country on that day. 

We can do this by summing up the cases across all of the states/provinces within each country and date.

```{r}


```

